---
title: "Vaccine Equity Analysis"
author: "C. Scott Smith"
date: "4/20/2021"
output: html_document
editor_options: 
chunk_output_type: console

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)

library(tidyverse)
library(dplyr) 
library(ggplot2)
library(lubridate)
library(zoo)
library(readxl)
library(sf)
library(DT)

```

## Background

### COVID-19 Burden in Chicago

```{r background information and maps}

adate <- as.character(Sys.Date())

Chicago_COVID19TestingLocations <- read_csv(file="https://data.cityofchicago.org/api/views/thdn-3grx/rows.csv?accessType=DOWNLOAD&bom=true&format=true")
Chicago_COVID19VaccinationLocations <- read_csv(file="https://data.cityofchicago.org/api/views/6q3z-9maq/rows.csv?accessType=DOWNLOAD")

```

### COVID-19 Burden in Chicago

## Research Questions

This research aims to respond to the following research questions: 

(1) Are there disparities in vaccination rates across Chicago communities with different race, ethnic and socioeconomic characteristics?

(2) If such disparities do exist, to what extent has Chicago's Protect Chicago Plus program mitigated them?

## Methodology

We employ a series of chi-square analyses to evaluate disparities in vaccination rates across communities categorized by  dominant race, ethnicity as well as COVID-19 burden. Interrupted time series analysis (ITS) is then used to evaluate the impact of the city's Protect Chicago Plus program on different measures of vaccination coverage and disparity, pre- and post-intervention.

### Data

To carry out the above analyses, we rely on publicly available datasets including the following:

(1) socioeconomic data including race, ethnicity and age composition from the US Census Bureau's American Community Survey (ACS);
(2) daily vaccination data from the City of Chicago's data portal; and 
(3) information from the City of Chicago and related media reports concerning the geographic scope and rollout of the city's Protect Chicago Plus program.


```{r import, format and summarize vaccination data from Chicago data portal}

# Download daily vaccination data by zip code from Chicago data portal
Chicago_COVID19Vaccinations_ByZCTA <- read_csv(file="https://data.cityofchicago.org/api/views/553k-3xzc/rows.csv?accessType=DOWNLOAD&bom=true&format=true")

# Rename columns, reformat date, create week field
Chicago_COVID19Vaccinations_ByZCTA <- Chicago_COVID19Vaccinations_ByZCTA %>% 
  rename("zcta"="Zip Code", 
         "date"="Date", 
         "doses_daily"="Total Doses - Daily", 
         "doses_cum"="Total Doses - Cumulative", 
         "fdose_daily"="1st Dose - Daily", 
         "fdose_cum"="1st Dose - Cumulative", 
         "fdose_pctpop"="1st Dose - Percent Population", 
         "com_daily"="Vaccine Series Completed - Daily", 
         "com_cum"="Vaccine Series Completed - Cumulative", 
         "com_pctpop"="Vaccine Series Completed  - Percent Population", 
         "pop"="Population", 
         "geometry"="ZIP Code Location") %>%
  select(-geometry) %>%
  mutate(date = as.Date(date, "%m/%d/%Y"),
         ymw = ceiling_date(date, "week"))

# Summarize by date
Chicago_COVID19Vaccinations_ByDate_Sum <- Chicago_COVID19Vaccinations_ByZCTA %>%
  filter(pop>0) %>% 
  group_by(date) %>% 
  arrange(date) %>% 
  summarise(doses = sum(doses_daily), 
            dosescm = sum(doses_cum), 
            fdoses = sum(fdose_daily),
            fdosescm = sum(fdose_cum), 
            sdoses = sum(com_daily), 
            sdosescm = sum(com_cum)) %>%
  slice(1:20) %>%
  datatable()

# Summarize by ZCTA
Chicago_COVID19Vaccinations_ByZCTA_Sum <- Chicago_COVID19Vaccinations_ByZCTA %>%
  filter(pop>0) %>% 
  group_by(zcta) %>% 
  summarise(doses = sum(doses_daily), 
            dosescm = sum(doses_cum), 
            fdoses = sum(fdose_daily),
            fdosescm = sum(fdose_cum), 
            sdoses = sum(com_daily), 
            sdosescm = sum(com_cum)) %>%
  mutate(ZCTA5=as.character(zcta))

```

``` {r Table}
Chicago_COVID19Vaccinations_ByDate_Sum
```

```{r import demographic data, calculate and summarize vaccination rates}

# Import previously downloaded 2019 ACS 5-year estimates demographic data by
# ZCTA

ZCTAs_ACS_Chicago <- read_csv("data/ZCTAs_ACS_Chicago.csv")

ZCTAs_ACS_Vac_geom <- Chicago_COVID19Vaccinations_ByZCTA %>%
  replace(., is.numeric(is.na(.)), "") %>%
  drop_na(zcta) %>%
  mutate(zcta = as.numeric(zcta)) %>%
  left_join(ZCTAs_ACS_Chicago, by=c("zcta"="ZCTA5")) %>%
  filter(Total>0) %>%
  select(-pop) %>%
  arrange(date,zcta) %>%
  group_by(zcta) %>%
  mutate(rec = 1,
         T = cumsum(rec),
         D = if_else(date>=as.Date("2021-02-05"),1,0),
         P = if_else(D>=1,cumsum(D),0),
         doses7d = rollmean(x=doses_daily, k=7, fill=0, align="right"), 
         dosert_totalpop = doses_cum/Total*100,
         dosert_65over = doses_cum/Pop65Over*100,
         dosert_18over = doses_cum/Pop18Over*100,
         dosert_16over = doses_cum/Pop16Over*100,
         dosert_essen = doses_cum/WESSEN*100,
         fdoses7d = rollmean(x=fdose_daily, k=7, fill=0, align="right"), 
         fdosert_totalpop = fdose_cum/Total*100,
         fdosert_65over = fdose_cum/Pop65Over*100,
         fdosert_18over = fdose_cum/Pop18Over*100,
         fdosert_16over = fdose_cum/Pop16Over*100,
         fdosert_essen = fdose_cum/WESSEN*100,
         sdoses7d = rollmean(x=com_daily, k=7, fill=0, align="right"), 
         sdosert_totalpop = com_cum/Total*100,
         sdosert_65over = com_cum/Pop65Over*100,
         sdosert_18over = com_cum/Pop18Over*100,
         sdosert_16over = com_cum/Pop16Over*100,
         sdosert_essen = com_cum/WESSEN*100)

```

``` {r format, summarize data for chi-square analysis}

# get date list from 2/1/2021 onward
aEndDateList <- ZCTAs_ACS_Vac_geom %>% 
  group_by(date) %>% 
  summarise() %>%
  drop_na()
aEndDateList <- aEndDateList[49:118,] # return subset of list
aEndDateList # view date list

ZCTA_select = st_read("layers/ZCTA_select.shp")
achisqtable_All <- readRDS("data/achisqtable_All.rds")
achisqtable_All <- achisqtable_All[c(), ]

ZCTAs_ACS_Vac_geom$NotVac_totpop <- ZCTAs_ACS_Vac_geom$Total-ZCTAs_ACS_Vac_geom$fdose_cum
ZCTAs_ACS_Vac_geom$NotVac_18over <- ZCTAs_ACS_Vac_geom$Pop18Over-ZCTAs_ACS_Vac_geom$fdose_cum
ZCTAs_ACS_Vac_geom$NotVac_16over <- ZCTAs_ACS_Vac_geom$Pop16Over-ZCTAs_ACS_Vac_geom$fdose_cum
ZCTAs_ACS_Vac_geom$NotVac_essen <- ZCTAs_ACS_Vac_geom$WESSEN-ZCTAs_ACS_Vac_geom$fdose_cum

# Run chi-square for all dates
for (i in 1:nrow(aEndDateList)) {
  ct_CountsbyCategory <- ZCTAs_ACS_Vac_geom %>%
    filter(date == aEndDateList$date[i]) %>%
    group_by(pcp) %>%
    drop_na() %>%
    summarize(Vaccinated = sum(fdose_cum),
              # NotVaccinated = sum(NotVac_totpop))
              NotVaccinated = sum(NotVac_18over))
  
  ct_CountsbyCategory[is.na(ct_CountsbyCategory)] <- 0
  ct_CountsbyCategory_t <- t(ct_CountsbyCategory[,-1])
  colnames(ct_CountsbyCategory_t) <- ct_CountsbyCategory$pcp
  
  chisq_Results <- chisq.test(ct_CountsbyCategory_t)
  achisqtable_obs <-  chisq_Results$observed
  achisqtable_exp <- round(chisq_Results$expected,2)
  achisqtable_exp <- achisqtable_exp[1,]
  achisqtable_obs <- achisqtable_obs[1,]
  achisqtable_cbind <- cbind(achisqtable_exp,achisqtable_obs)
  achisqtable_cbind <- as.data.frame.matrix(achisqtable_cbind)
  achisqtable_cbind$date <- aEndDateList$date[i]
  achisqtable_cbind$PValue <- chisq_Results$p.value
  achisqtable_cbind$XSqrd <- chisq_Results$statistic
  achisqtable_cbind$df <- chisq_Results$parameter
  achisqtable_cbind <- cbind(rownames(achisqtable_cbind), data.frame(achisqtable_cbind, row.names=NULL))
  achisqtable_All <- rbind(achisqtable_cbind,achisqtable_All)
}

achisqtable_All <- achisqtable_All %>% 
  rename("Vaccinated"=`rownames(achisqtable_cbind)`,
         "Expected"='achisqtable_exp',
         "Observed"='achisqtable_obs')

achisqtable_All$Residual <- achisqtable_All$Observed - achisqtable_All$Expected


# create residual plot
ggplot(achisqtable_All) + geom_point(aes(x=date,y=Residual, color = Vaccinated), shape=1, size = 3, stroke=2) + geom_line(aes(x=date,y=Residual, color = Vaccinated), size=1.5) + scale_color_manual(values=c('#A6758D','#8DB1D5','#FFC000','#FBA2A2')) + theme(legend.position="top", legend.title = element_blank(), axis.title.x = element_blank(), legend.key=element_blank(), axis.ticks=element_line(size=1), text = element_text(family="arial", face="bold", size=18), axis.text.x = element_text(angle = 90)) 

# clear subsequent chi-square analyses
achisqtable_All <- readRDS("data/achisqtable_All.rds")
achisqtable_All <- achisqtable_All[c(), ]


```


``` {r format, summarize data for interrupted time series analysis}

ZCTAs_ACS_Vac_ByPCP_geom <- ZCTAs_ACS_Vac_geom %>%
  drop_na(zcta) %>%
  filter(Total>0) %>%
  arrange(pcp, date) %>%
  group_by(pcp, date) %>%
  summarize(rec = 1, 
            doses_cum = sum(doses_cum),
            Total = sum(Total)) %>%
  mutate(Y = doses_cum/Total*100, 
         D=if_else(date>=as.Date("2021-02-05"),1,0),
         P = if_else(D>=1,cumsum(D),0),
         T = cumsum(rec))

ZCTAs_ACS_Vac_ByPCP_geom <- ZCTAs_ACS_Vac_geom %>%
  drop_na() %>%
  select(pcp, date, doses_cum, Total) %>%
  group_by(pcp) %>%
  summarize(rec = 1,
            Y = sum(doses_cum)/sum(Total)*100,
            D = if_else(date>=as.Date("2021-02-05"),1,0),
            P = if_else(D>=1,cumsum(D),0))

ZCTAs_ACS_Vac_geom %>% select(date,D,P) %>% group_by(D) %>% summarise(n())

```

## Resources

https://ds4ps.org/pe4ps-textbook/docs/p-020-time-series.html
https://ds4ps.org/pe4ps-textbook/docs/p-020-time-series.html
https://online.stat.psu.edu/stat462/node/188/
https://www-sciencedirect-com.ezproxy.depaul.edu/science/article/pii/S0277953602005026
https://worldwidescience.org/topicpages/v/vaccination+odds+ratio.html
https://www.quora.com/What-is-an-equivalent-of-the-chi-square-test-for-time-series-data
