---
title: "Vaccine Equity Analysis"
author: "C. Scott Smith"
date: "4/20/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr) 
library(ggplot2)
library(lubridate)
library(zoo)
library(readxl)

```

## Background


## Research Questions

To what extent do neighborhoods 

## Methodology

We implement an interrupted time series analysis to evaluate the impact of a Protect Chicago Plus vaccination over time. We compare longitudinal vaccination data across two groups of neighborhoods across the city, comparing rates throughout a three month period. We wish to understand how and if the share vaccination rates changed after an intervention, a policy, or a program that was implemented within the population.

### Data

This analysis relies on a combination of publicly available datasets including the following:

(1) The US Census Bureau's American Community Survey
(2) TIGER data, ZCTA boundaries
(3) Chicago data portal daily vaccination data
(4) Information from the Protect Chicago Plus program including participating zip codes 


```{r import, format and summarize vaccination data from Chicago data portal}

# Download daily vaccination data by zip code from Chicago data portal
Chicago_COVID19Vaccinations_ByZCTA <- read_csv(file="https://data.cityofchicago.org/api/views/553k-3xzc/rows.csv?accessType=DOWNLOAD&bom=true&format=true")

# Rename columns, reformat date, create week field
Chicago_COVID19Vaccinations_ByZCTA <- Chicago_COVID19Vaccinations_ByZCTA %>% 
  rename("zcta"="Zip Code", 
         "date"="Date", 
         "doses_daily"="Total Doses - Daily", 
         "doses_cum"="Total Doses - Cumulative", 
         "fdose_daily"="1st Dose - Daily", 
         "fdose_cum"="1st Dose - Cumulative", 
         "fdose_pctpop"="1st Dose - Percent Population", 
         "com_daily"="Vaccine Series Completed - Daily", 
         "com_cum"="Vaccine Series Completed - Cumulative", 
         "com_pctpop"="Vaccine Series Completed  - Percent Population", 
         "pop"="Population", 
         "geometry"="ZIP Code Location") %>%
  select(-geometry) %>%
  mutate(date = as.Date(date, "%m/%d/%Y"),
         ymw = ceiling_date(date, "week"))

# Summarize by date
Chicago_COVID19Vaccinations_ByDate_Sum <- Chicago_COVID19Vaccinations_ByZCTA %>%
  filter(pop>0) %>% 
  group_by(date) %>% 
  arrange(date) %>% 
  summarise(doses = sum(doses_daily), 
            dosescm = sum(doses_cum), 
            fdoses = sum(fdose_daily),
            fdosescm = sum(fdose_cum), 
            sdoses = sum(com_daily), 
            sdosescm = sum(com_cum))

# Summarize by ZCTA
Chicago_COVID19Vaccinations_ByZCTA_Sum <- Chicago_COVID19Vaccinations_ByZCTA %>%
  filter(pop>0) %>% 
  group_by(zcta) %>% 
  summarise(doses = sum(doses_daily), 
            dosescm = sum(doses_cum), 
            fdoses = sum(fdose_daily),
            fdosescm = sum(fdose_cum), 
            sdoses = sum(com_daily), 
            sdosescm = sum(com_cum)) %>%
  mutate(ZCTA5=as.character(zcta))

```


```{r import demographic data}

# Import previously downloaded 2019 ACS 5-year estimates demographic data by
# ZCTA

ZCTAs_ACS_Chicago <- read_csv("data/ZCTAs_ACS_Chicago.csv")

ZCTAs_ACS_Vac_geom <- Chicago_COVID19Vaccinations_ByZCTA %>%
  replace(., is.numeric(is.na(.)), "") %>%
  drop_na(zcta) %>%
  left_join(ZCTAs_ACS_Chicago, by=c("zcta"="ZCTA5")) %>%
  filter(Total>0) %>%
  select(-pop) %>%
  arrange(date,zcta) %>%
  group_by(zcta) %>%
  mutate(rec = 1,
         T = cumsum(rec),
         D = if_else(date>=as.Date("2021-02-05"),1,0),
         P = if_else(D>=1,cumsum(D),0),
         doses7d = rollmean(x=doses_daily, k=7, fill=0, align="right"), 
         dosert_totalpop = doses_cum/Total*100,
         dosert_65over = doses_cum/Pop65Over*100,
         dosert_18over = doses_cum/Pop18Over*100,
         dosert_16over = doses_cum/Pop16Over*100,
         dosert_essen = doses_cum/WESSEN*100,
         fdoses7d = rollmean(x=fdose_daily, k=7, fill=0, align="right"), 
         fdosert_totalpop = fdose_cum/Total*100,
         fdosert_65over = fdose_cum/Pop65Over*100,
         fdosert_18over = fdose_cum/Pop18Over*100,
         fdosert_16over = fdose_cum/Pop16Over*100,
         fdosert_essen = fdose_cum/WESSEN*100,
         sdoses7d = rollmean(x=com_daily, k=7, fill=0, align="right"), 
         sdosert_totalpop = com_cum/Total*100,
         sdosert_65over = com_cum/Pop65Over*100,
         sdosert_18over = com_cum/Pop18Over*100,
         sdosert_16over = com_cum/Pop16Over*100,
         sdosert_essen = com_cum/WESSEN*100)

ZCTAs_ACS_Vac_ByPCP_geom <- ZCTAs_ACS_Vac_geom %>%
  drop_na(zcta) %>%
  filter(Total>0) %>%
  arrange(pcp, date) %>%
  group_by(pcp, date) %>%
  summarize(rec = 1, 
            doses_cum = sum(doses_cum),
            Total = sum(Total)) %>%
  mutate(Y = doses_cum/Total*100, 
         D=if_else(date>=as.Date("2021-02-05"),1,0),
         P = if_else(D>=1,cumsum(D),0),
         T = cumsum(rec))

ZCTAs_ACS_Vac_ByPCP_geom <- ZCTAs_ACS_Vac_geom %>%
  drop_na() %>%
  select(pcp, date, doses_cum, Total) %>%
  group_by(pcp) %>%
  summarize(rec = 1,
            Y = sum(doses_cum)/sum(Total)*100,
            D = if_else(date>=as.Date("2021-02-05"),1,0),
            P = if_else(D>=1,cumsum(D),0))

ZCTAs_ACS_Vac_geom %>% select(date,D,P) %>% group_by(D) %>% summarise(n())

ZCTAs_ACS_Vac_geom$NotVac_totpop <- ZCTAs_ACS_Vac_geom$Total-ZCTAs_ACS_Vac_geom$fdose_cum
ZCTAs_ACS_Vac_geom$NotVac_18over <- ZCTAs_ACS_Vac_geom$Pop18Over-ZCTAs_ACS_Vac_geom$fdose_cum
ZCTAs_ACS_Vac_geom$NotVac_16over <- ZCTAs_ACS_Vac_geom$Pop16Over-ZCTAs_ACS_Vac_geom$fdose_cum
ZCTAs_ACS_Vac_geom$NotVac_essen <- ZCTAs_ACS_Vac_geom$WESSEN-ZCTAs_ACS_Vac_geom$fdose_cum

```

## Resources

https://ds4ps.org/pe4ps-textbook/docs/p-020-time-series.html
https://ds4ps.org/pe4ps-textbook/docs/p-020-time-series.html
https://online.stat.psu.edu/stat462/node/188/
https://www-sciencedirect-com.ezproxy.depaul.edu/science/article/pii/S0277953602005026
https://worldwidescience.org/topicpages/v/vaccination+odds+ratio.html
https://www.quora.com/What-is-an-equivalent-of-the-chi-square-test-for-time-series-data
